### 2016-02-12 prepReport/earlyWork.tex
Applied all improvements suggested

\chapter{Early Work}\label{chap:earlyWork}
In order to better undesrtand LSTM,  gain some useful insight on their operation, and to have a software implementation, that is not part of any framework, with which I can validate in the future my FPGA implementation, I have developed an fully functional LSTM Layer in Python and Numpy, that still uses BPTT as a training algorithm (the governing equations are in appendix A of~\cite{Greff15}). The idea is to adapt it to use SPSA in the beginning of my Master's Thesis. The code is present in

>>>>

\chapter{Early Work}\label{chap:earlyWork}
In order to better undesrtand LSTM,  gain some useful insight on their operation, and to have a software implementation, that is not part of any framework, with which I can validate in the future my FPGA implementation, I have developed a fully functional LSTM Layer in Python and Numpy, that still uses BPTT as a training algorithm (the governing equations are in appendix A of~\cite{Greff15}). The idea is to adapt it to use SPSA in the beginning of my Master's work. The code is present in

### 2016-02-12 prepReport/theorBack.tex
Applied all improvements suggested

    \caption[Representation of a Recurrent Neuron (Figure~\ref{fig:recneuron}) and a Recurrent Neuron unfolded through time (Figure~\ref{recneuron_unf})]{In Figure~\ref{fig:recneuron}, a recurrent neuron is depicted, where the output is fed back to the input, weighted by $\theta_h$. Figure~\ref{fig:recneuron_unf}, an unfolding through time of that same neuron is performed. The basic idea is that feeding the output back to the input is similar to feeding the output to the input of the neuron \emph{at a previous time step}: this way, we linearize the structure. Considering a training epoch of $T$ samples, corresponds to having $T$ unfolded neurons/layers}
	\label{fig:recurrneuron}
\end{figure}

@@

\subsubsection{Structure, Operation and Equations}\label{sec:struct_lstm}
A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As we can see from the picture, we still have the recurrent connections from the regular RNNs, but now there are multiple entry points that control the flow of information through the network. Although omitted from the picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLstm}. The main components, their role and relevance, are explained as follows

@@

\begin{align}\label{eq:equationsLSTM}
    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \\
    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \\
    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \\
    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \\
    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \\
    \mb{c}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) 
\end{align}
where $\odot$ is the Hadamard multiplication. The $i$-th element of the previous vectors in bold corresponds to the value of the gate of the $i$-th neuron of the layer, which is a very convenient and compact representation of the whole layer. Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that precedes this), we see that the input weight matrices $\mb{W}_*$ and the  recurrent weight matrices $\mb{R}_*$ all have size $N \times M$, and that the bias weight matrices $\mb{b}_*$, the peephole weight matrices $\mb{p}_*$ and the matrices $\mb{z}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$. Although useful for a high-level description in a programming language, the matrix representation  may not be suitable for a direct HDL port. In that case, in order to represent the operation of the $i$-\textit{single neuron}, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single scalar output, and we only use the appropriate row in the weight matrices $\mb{W}_{*}$, $\mb{R}_{*}$ and the remaining.

@@

\begin{equation}\label{eq:taylor}
    \Delta w_i^{(t)} = s_i^{(t)} (\mb{s}^{(t)})^T \frac{\partial J(\mb{w}^{(t)})}{\partial \mb{w}} +
                        \frac{c s_i^{(t)}}{2} (\mb{s}^{(t)})^T \frac{\partial^2 J(\mb{w}^{(t)})}{\partial \mb{w}^2}  (\mb{s}^{(t)}) 
\end{equation}
If the random sign vector $\mb{s}^{(t)}$ is chosen carefully so that the expected value of the vector at the $i$-th position is zero, $E(s_i^{(t)})$, and the signs of two different components $i$ and $j$, at different time-steps, are independent (i.e. $E(s_i^{(t_1)}s_j^{(t_2)}) = \delta_{ij}\delta_{t_1 t_2}$, equation~\ref{eq:taylor} reduces to

@@

\section{Proposed Solution}\label{sec:propSol}
Given the previous discussion of how LSTM neurons work and their benefits when compared to other RNN architectures, this work aims to conceive and design an efficient dedicated FPGA implementation with on-chip learning that, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or of a framework (see Section~\re (see Section~\ref{sec:wp_sh}).

>>>>

    \caption[Representation of a Recurrent Neuron (Figure~\ref{fig:recneuron}) and a Recurrent Neuron unfolded through time (Figure~\ref{fig:recneuron_unf})]{In Figure~\ref{fig:recneuron}, a recurrent neuron is depicted, where the output is fed back to the input, weighted by $\theta_h$. Figure~\ref{fig:recneuron_unf}, an unfolding through time of that same neuron is performed. The basic idea is that feeding the output back to the input is similar to feeding the output to the input of the neuron \emph{at a previous time step}: this way, we linearize the structure. Considering a training epoch of $T$ samples, corresponds to having $T$ unfolded neurons/layers}
	\label{fig:recurrneuron}
\end{figure}

@@

\subsubsection{Structure, Operation and Equations}\label{sec:struct_lstm}
A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As we can see from the picture, we still have the recurrent connections from the regular RNNs, but now there are multiple entry points that control the flow of information through the network. Although omitted from the picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLSTM}. The main components, their role and relevance, are explained as follows

@@

\begin{align}
    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \notag\\
    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \notag\\
    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \notag\\
    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \notag \\
    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \notag\\
    \mb{c}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
\end{align}
where $\odot$ is the Hadamard multiplication. The $i$-th element of the previous vectors in bold corresponds to the value of the gate of the $i$-th neuron of the layer, which is a very convenient and compact representation of the whole layer. Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that precedes this), we see that the input weight matrices $\mb{W}_*$ and the  recurrent weight matrices $\mb{R}_*$ all have size $N \times M$, and that the bias weight matrices $\mb{b}_*$, the peephole weight matrices $\mb{p}_*$ and the matrices $\mb{z}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$. Although useful for a high-level description in a programming language, the matrix representation  may not be suitable for a direct HDL port. In that case, in order to represent the operation of the $i$-\textit{single neuron}, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single scalar output, and we only use the appropriate row in the weight matrices $\mb{W}_{*}$, $\mb{R}_{*}$ and the remaining.

@@

\begin{equation}\label{eq:taylor}
    \Delta w_i^{(t)} = s_i^{(t)} (\mb{s}^{(t)})^T \frac{\partial J(\mb{w}^{(t)})}{\partial \mb{w}} +
                        \frac{c s_i^{(t)}}{2} (\mb{s}^{(t)})^T \frac{\partial^2 J(\mb{w}^{(t)})}{\partial \mb{w}^2}  (\mb{s}^{(t)}) 
\end{equation}
If the random sign vector $\mb{s}^{(t)}$ is chosen carefully so that the expected value of the vector at the $i$-th position is zero, $E(s_i^{(t)})$, and the signs of two different components $i$ and $j$, at different time-steps, are independent (i.e. $E(s_i^{(t_1)}s_j^{(t_2)}) = \delta_{ij}\delta_{t_1 t_2}$, where $\delta$ is the Kronecker fuction), taking the \emph{Expected Value} of equation~\ref{eq:taylor} yields

@@

\section{Proposed Solution}\label{sec:propSol}
Given the previous discussion of how LSTM neurons work and their benefits when compared to other RNN architectures, this work aims to conceive and design an efficient dedicated FPGA implementation with on-chip learning that, surpasses the training/testing time of an equivalent CPU implementation in a general high-level language or of a framework (see Section~\ref{sec:wp_sh}).

### 2016-02-12 prepReport/workPlan.tex
Applied all improvements suggested

\section{Approach and  Main Tasks}
Dwight Eisenhower, United States' President from 1953 to 1961 and Army Officer, said that `no battle was ever won according to plan, but no battle was ever won without one'. Hence, although this initial planning draft might not be religiously followed in all its extent, it is important to break down the work in \textbf{tasks} so that I and my supervisors can keep track of what has to be done, and establishing deadlines for the small achievements that, together, will build up to the successful completion of the proposed work of this thesis. The detailed explanation of each task is provided in Section~\ref{sec:wp_tasks_td} and the \textbf{schedule} for them, where their approximate begin and end weeks are marked, is represented by a \emph{Gantt Chart} in Section~\ref{sec:wp_tasks_gantt}.

\subsection{Task Description}\label{sec:wp_tasks_td}
The work will be partitioned in \emph{five} large groups of tasks, and each of these groups will be further partitioned in smaller subtasks which are the actual assignments that I will be completing. We can think of the group of task as a sort of \emph{logical} grouping of the subtasks, according both to their relatedness and also to their precedence relationships. The task groups, and their related subtasks, are as follows

@@

        \begin{itemize}
            \item \textbf{T1.1 -- Complete Python Class with SPSA} -- here I will finish the Python modules that I have already developed as a preparation for this work, whose details are presented in Section~\ref{chap:earlyWork}, by adding the SPSA training algorithm, as opposed to the current Backpropagation Through Time version.

            \item \textbf{T1.2 -- Reproduce the Benchmark Results with the Python Class} -- after developing and testing the Python Class with SPSA, I will try to reproduce with it the results of~\cite{Tavear13}, in order for me to have a high-level model whose performance is comparable to~\cite{Tavear13, Hochreiter07}, and thus provide a good performance baseline for this and other future benchmark tests that I would like to run in order to compare my FPGA solution with a software solution.

@@

        \begin{itemize}
            \item \textbf{T2.1 -- Design Planning} -- before performing and HDL coding, it is favourable to first assemble the system design strategy in paper. This is where, together with my supervisor, I will outline the internal organization of my system, and make the fundamental design decisions. Of course, this task will prolong during all the \textbf{T2} extent, since some design changes to the original plan will have to be performed as I find \emph{bugs} in my early Verilog implementations that force a design correction.

            \item \textbf{T2.2 -- Design SPSA Module} -- this task comprises the Verilog description of the module that will update the network weights using SPSA, and also the development of a testbench to validate my description. Furthermore, I will compare the output of this model with the golden output produced by a standalone SPSA Python Class previously developed in \textbf{T1}.
            \item \textbf{T2.3 -- Design LSTM Neuron/Layer Module} -- similar to the previous task, but now I am aiming to implement the actual LSTM Neuron or a layer of LSTM Neurons. This is a design decision that will need to be set in \textbf{T2.1}, to assess whether it is justifiable to have dedicated multiplication units assigned to each neuron, or we can perform matrix-like batch multiplications and treat the LSTM as a batch of Neurons, i.e. as a layer. Like in the previous task, a testbench will also be developed, and the module validate both with it and the Python golden output.

@@

    \item \textbf{T3 -- FPGA Synthesis of the Verilog Code} -- after having Verilog code that passed all the validation tests and whose correct functionality is, therefore, asserted, I will synthesize the Verilog modules to FPGA RTL blocks, and perform a new simulation on these RTL syntesized code, that unlike the previous behavioural simulation, now has timing information resulting from the FPGA block mapping, and the place and routing of those mapped blocks.

@@

    \item \textbf{T5 -- Write the Dissertation Final Report} -- while all the stages of work are developing, this task will obviously be active, by documenting all the design decisions, milestones achieved and logging all the results that come from the implementation. A couple of weeks before the deadline, this will be the only active task, since producing a clear and well-written document requires concentration and undivided attention.

@@

\subsection{Hardware Resources}
The main hardware resource that I will use for my solution is the \textbf{Xilinx\textregistered~ Zedboard} (Figure~\ref{fig:zedboard}, whose core is a Zynq-7000 System-on-Chip that features both an FPGA core and an ARM \textregistered~ Cortex\texttrademark~-A9 processor. It is an FPGA development board that features several I/O interfaces such as Ethernet, HDMI output, VGA output, Audio IOs, USB and UART, among other capabilities, making it easy to prototype additional features to the core design (i.e. for building the note transcription system of~\cite{Bock12}, and to make it operate in real-time, I would need an audio input, which I can easily get from the Audio IO, without investing time in connecting the codec and programming the FPGA to interface it).

@@

To write the final Dissertation report, I will use \LaTeX (this report itself was also written in \LaTeX), and mostly use Tikz, Inkscape and Matplotlib to produce figures and drawings. Once again, all of this is free to use.

>>>>

\section{Approach and  Main Tasks}
Dwight Eisenhower, United States' President from 1953 to 1961 and Army Officer, said that `no battle was ever won according to plan, but no battle was ever won without one'. Hence, although this initial planning draft might not be religiously followed in all its extent, it is important to break down the work in \textbf{tasks} so that I and my supervisors can keep track of what has to be done, and establishing deadlines for the small achievements that, together, will build up to the successful completion of the proposed work of this thesis. The detailed explanation of each task is provided in Section~\ref{sec:wp_tasks_td} and the \textbf{schedule} for them, where their estimated begin and end weeks are marked, is represented by a \emph{Gantt Chart} in Section~\ref{sec:wp_tasks_gantt}.

\subsection{Task Description}\label{sec:wp_tasks_td}
The work will be partitioned in \emph{five} large groups of tasks, and each of these groups will be further partitioned in smaller subtasks which are the actual assignments that I will be completing. We can think of the group of task as a sort of \emph{logical} grouping of the subtasks, according both to their relatedness and also to their precedence relationships. The task groups, and their related subtasks, are as follows:

@@

        \begin{itemize}
            \item \textbf{T1.1 -- Complete Python Implementation with SPSA} -- here I will finish the Python modules that I have already developed as a preparation for this work, whose details are presented in Section~\ref{chap:earlyWork}, by adding the SPSA training algorithm, as opposed to the current Backpropagation Through Time version.

            \item \textbf{T1.2 -- Reproduce the Benchmark Results with the Python Implementation} -- after developing and testing the Python Class with SPSA, I will try to reproduce with it the results of~\cite{Tavear13}, in order for me to have a high-level model whose performance is comparable to~\cite{Tavear13, Hochreiter07}, and thus provide a good performance baseline for this and other future benchmark tests that I would like to run in order to compare my FPGA solution with a software solution.

@@

        \begin{itemize}
            \item \textbf{T2.1 -- Design Planning} -- before performing and HDL coding, it is desirable first assemble the system design strategy in paper. This is where, together with my supervisor, I will outline the internal organization of my system, and make the fundamental design decisions. Of course, this task will continue during the entire of \textbf{T2} extent, since some design changes to the original plan will have to be performed as I find \emph{bugs} in my early Verilog implementations that force a design correction.

            \item \textbf{T2.2 -- Design SPSA Module} -- this task comprises the Verilog description of the module that will update the network weights using SPSA, and also the development of a testbench to validate my description. Furthermore, I will compare the output of this model with the golden output produced by a standalone SPSA Python Class previously developed in \textbf{T1}.

            \item \textbf{T2.3 -- Design LSTM Neuron/Layer Module} -- similarly to the previous task, but now I am aiming to implement the actual LSTM Neuron or a layer of LSTM Neurons. This is a design decision that will need to be set in \textbf{T2.1}, to assess whether it is justifiable to have dedicated multiplication units assigned to each neuron, or we can perform matrix-like batch multiplications and treat the LSTM as a batch of Neurons, i.e. as a layer. Like in the previous task, a testbench will also be developed, and the module validate both with it and the Python golden output.

@@

    \item \textbf{T3 -- FPGA Synthesis of the Verilog Code} -- after having Verilog code that passed all the validation tests and whose correct functionality is, therefore, ensured, I will synthesize the Verilog modules to FPGA RTL blocks, and perform a new simulation on these RTL syntesized code, that unlike the previous behavioural simulation, now has timing information resulting from the FPGA block mapping, and the place and routing of those mapped blocks.

@@

    \item \textbf{T5 -- Write the Dissertation Final Report} -- while all the stages of work are progressing, this task will obviously be active, by documenting all the design decisions, milestones achieved and logging all the results that come from the implementation. A couple of weeks before the deadline, this will be the only active task, since producing a clear and well-written document requires concentration and undivided attention.

@@

\subsection{Hardware Resources}
The main hardware resource that I will use for my solution is the \textbf{Xilinx\textregistered~ Zedboard} (Figure~\ref{fig:zedboard}), whose core is a Zynq-7000 System-on-Chip that features both an FPGA core and two ARM \textregistered~ Cortex\texttrademark~-A9 processors. It is an FPGA development board that features several I/O interfaces such as Ethernet, HDMI output, VGA output, Audio IOs, USB and UART, among other capabilities, making it easy to prototype additional features to the core design (i.e. for building the note transcription system of~\cite{Bock12}, and to make it operate in real-time, I would need an audio input, which I can easily get from the Audio IO, without investing time in connecting the codec and programming the FPGA to interface it).

@@

To write the final Dissertation report, I will use \LaTeX~(this report itself was also written in \LaTeX), and mostly use Tikz, Inkscape and Matplotlib to produce figures and drawings. Once again, all of these are free to use.
